---
title: "Powerlifting EDA - SQL"
author: "Samad Patel"
date: "9/15/2018"
output: pdf_document
---

```{r Libraries, Connection, echo = FALSE, warning=FALSE, message=FALSE}
# Libraries
library(readr)
library(dplyr)
library(ggplot2)
library(ggthemes)
library(lubridate)
library(gridExtra)
library(RMySQL)
con <- dbConnect(MySQL(), 
                 user = 'root', 
                 password = 'spronwhikkle', 
                 host = 'localhost', 
                 dbname = 'powerlifting')
```

# Investigation By Year
# Average Age 
We will investigate from 1998 onward, as we have complete data for every year from that point on. 
```{r, echo = FALSE, warning = FALSE, message=FALSE}
# Write our sql query
age_query <- '
SELECT YEAR(meets.Date) AS Year, AVG(lifts2.Age) AS Avg_Age FROM lifts2
LEFT JOIN meets USING(MeetID)
GROUP BY Year
HAVING Year >= 1998
ORDER BY Year;
'
# Generate dataframe with query
age <- dbGetQuery(con, age_query)

# Create plot 
ggplot(age) + geom_line(aes(x=Year, y = Avg_Age, colour = 'blue'), size = 1.3) + 
   theme_fivethirtyeight() + scale_color_fivethirtyeight(guide = FALSE) + 
   scale_x_continuous(breaks = seq(1982, 2018, by = 6)) + 
   theme(plot.title = element_text(hjust = .5), axis.title = element_text()) +
      scale_x_continuous(breaks = seq(1998, 2018, 2), name = seq(1998, 2018, 2)) + 
      ggtitle('Average Age of Competitors Over Time') + ylab('Average Age') + 
      xlab('Year')
```

There is no clear trend overall, however it appears as though the average age has been decreasing in the last decade. 


## Sexes
First, let's observe the number of women in Powerlifting over time. 
```{r, echo = FALSE, message=FALSE, warning=FALSE}
# Write SQL Query
women_query <- '
SELECT YEAR(m.Date) AS Year, 
       SUM(CASE WHEN l.Sex = \'F\' THEN 1 ELSE 0 END) AS Num_Women 
FROM lifts2 AS l
LEFT JOIN meets AS m USING(MeetID)
GROUP BY Year
ORDER BY Year; 
'

# Generate df using query
women <- dbGetQuery(con, women_query)
# Filter 1997 onwards 
ggplot(women %>% filter(Year >= 1997 & Year <=2017)) + 
   geom_bar(stat = 'identity', aes(x=Year, y=Num_Women, fill = 'blue')) + 
   theme_fivethirtyeight() + scale_fill_fivethirtyeight(guide = FALSE) + 
   scale_x_continuous(breaks = seq(1997, 2017, 2)) + 
   scale_y_continuous(breaks = seq(0, 20000, 2500)) + 
   theme(axis.title = element_text(), plot.subtitle = element_text(hjust =.5), 
         axis.text.x = element_text(angle = 45), plot.title = element_text(hjust=.5)) + 
   ggtitle('Women in PowerLifting Meets', subtitle = 'Per Year') + 
   ylab('Number of Women')
```

Now let's observe the proportion of women to men in the last decade. 
```{r, echo = FALSE, message=FALSE, warning=FALSE}
both_genders_query <- '
SELECT Sex, COUNT(Sex), YEAR(meets.Date) as Year FROM lifts2
INNER JOIN meets USING(MeetID)
GROUP BY Year, Sex
HAVING Year >= 1996
ORDER BY Year;' 

both_genders <- dbGetQuery(con, both_genders_query)
ggplot(both_genders %>% filter(Year <= 2017)) + 
   geom_bar(stat = 'identity', aes(x = Year, y = `COUNT(Sex)`, fill = Sex)) +
   theme_fivethirtyeight() + scale_fill_manual(values = c('#FF2700', "#008FD5")) + 
   scale_x_continuous(breaks = seq(1996, 2017, 1)) + 
   theme(axis.title = element_text(), plot.subtitle = element_text(hjust =.5), 
         axis.text.x = element_text(angle = 45), plot.title = element_text(hjust=.5),
         legend.position = 'right', legend.direction = 'vertical') + 
   ggtitle('Men and Women in PowerLifting Meets', subtitle = 'Per Year') + 
   ylab('Number of Competitors')
```

We can also view the explicit ratio. 
```{r, echo = FALSE, message=FALSE, warning=FALSE}
ratio_query <- '
SELECT YEAR(meets.Date) AS Year, 
	   SUM(CASE WHEN lifts2.Sex = \'F\' THEN 1 ELSE 0 END) / 
      SUM(CASE WHEN lifts2.Sex = \'M\' THEN 1 ELSE 0 END) AS M2F_RATIO
FROM lifts2
LEFT JOIN meets ON lifts2.MeetID = meets.MeetID
GROUP BY Year
HAVING Year >= 1998 AND Year < 2018
ORDER BY Year;'

ratios <- dbGetQuery(con, ratio_query)
ggplot(ratios) + geom_line(aes(x=Year, y = M2F_RATIO), size = 1.3, colour = '#FF2700') + 
   theme_fivethirtyeight() + 
   scale_x_continuous(breaks = seq(1996, 2017, by = 1)) + 
   #scale_y_continuous(breaks = seq(1,8,1)) + 
   theme(axis.title = element_text(), plot.subtitle = element_text(hjust =.5), 
         axis.text.x = element_text(angle = 45), plot.title = element_text(hjust=.5),
         legend.position = 'right', legend.direction = 'vertical') +  
      ggtitle('Number of Females Per One Male') + ylab('Ratio') + 
      xlab('Year')
```

We can clearly observe that more and more women have been competing in powerlifitng over time, with the greatest changes and consistency in trend happening in the last decade. 

# Bench, Squat, and Deadlift Over Time
```{r, echo = FALSE, message=FALSE, warning=FALSE}
lifts_query <- '
SELECT MAX(lifts2.BestBenchKg) * 2.20462 AS Bench, 
       MAX(lifts2.BestSquatKg) * 2.20462 AS Squat, 
       MAX(lifts2.BestDeadliftKg) * 2.20462 AS Deadlift,
       YEAR(meets.Date) AS Year
FROM lifts2
INNER JOIN meets USING(MeetID)
GROUP BY Year
HAVING YEAR < 2018
ORDER BY Year;
'

lifts <- dbGetQuery(con, lifts_query)
ggplot(lifts) + geom_line(aes(Year, Bench, colour = 'Bench')) + 
   geom_line(aes(Year, Squat, color = 'Squat')) + 
   geom_line(aes(Year, Deadlift, color = 'Deadlift')) + 
   scale_x_continuous(breaks = seq(min(lifts$Year), max(lifts$Year), 5)) + 
   scale_color_fivethirtyeight(name = 'Lift') + theme_fivethirtyeight() + 
   theme(axis.title = element_text(), plot.subtitle = element_text(hjust =.5), 
         axis.text.x = element_text(angle = 45), plot.title = element_text(hjust=.5),
         legend.position = 'right', legend.direction = 'vertical', 
         legend.title = element_text(hjust = .5)) + 
   ggtitle('Max Lifts Over Time') + 
   ylab('Weight (lbs)') 
```
The last few years appear a bit suspicious to me, given that Bench is higher than Deadlift by quite a margin. But besides that, it's quite remarkable how similarly they all change together. That makes me interested in what the correlation matrix might look like. 

## Correlation Matrix
```{r, echo = FALSE, message=FALSE, warning=FALSE}
library(corrplot)
corr_mat <- cor(lifts[,1:3])

corrplot(corr_mat, method = 'color', type = 'lower', addCoef.col = 'white', 
         tl.col="black", tl.srt=0, diag = FALSE, title = 'Correlation of Lifts')
```


# Maps
We can use data from the Meet table to create some maps. 

## Map of Average Wilks Across the Continental US
```{r, echo = FALSE, warning=FALSE}
wilks_query <- '
SELECT meets.MeetState, AVG(lifts2.Wilks) AS Wilks FROM lifts2
LEFT JOIN meets USING(MeetID)
WHERE meets.MeetCountry = \'USA\'
GROUP BY meets.MeetState
ORDER BY meets.MeetState; 
'
wilks <- dbGetQuery(con, wilks_query)
# Change DC to VA
wilks[wilks$MeetState == 'DC', 'MeetState'] <- 'VA'
# Remove Alaska, Hawaii and Blank state 
wilks <- wilks %>% filter(MeetState != '' & MeetState != 'AK' & MeetState != 'HI')
# Change abbreviation to name
wilks$MeetState <- state.name[match(wilks$MeetState, state.abb)] %>% tolower()
# Change name of meetstate to region
colnames(wilks)[1] <- 'region'
# Create map data
usa <- map_data('state')
# Map to ggplot
map <- ggplot(data = usa) + geom_map(map = usa, aes(long, lat, map_id = region))
map + geom_map(data = wilks, map = usa, aes(fill = Wilks, map_id = region)) + 
   scale_fill_gradient(low = 'white', high = 'cyan4') + 
   theme(axis.title = element_blank(), panel.grid = element_blank(), 
         panel.background = element_blank(), axis.ticks = element_blank(), 
         axis.text = element_blank(), plot.title = element_text(hjust = .5, size = 16)) + 
   ggtitle('Average Wilks Score By State')
```

## Number of Meets Relative to Population 
Here we will invstigate how many meets there are in each state relative to the State's population. We will make use of the build in dataset state.x77, which contains information concerning state populations. This information is from 1977, so it's not up to date. 

```{r, echo = FALSE, warning=FALSE, message=FALSE}
num_meets_query <- '
SELECT meets.MeetState, COUNT(lifts2.MeetID) AS Num_Meets FROM lifts2
LEFT JOIN meets USING(MeetID) 
WHERE meets.MeetCountry = \'USA\' AND meets.MeetState != \'\'
GROUP BY meets.MeetState
ORDER BY Num_Meets;'

num_meets <- dbGetQuery(con, num_meets_query)
# Add DC meets to VA meets
num_meets[num_meets$MeetState == 'VA', 2] <- num_meets[num_meets$MeetState == 'VA', 2] + 
                                             num_meets[num_meets$MeetState == 'DC', 2]
# Remove DC, HI, AK
num_meets <- num_meets %>% filter(MeetState != 'DC' & MeetState != 'HI' & MeetState != 'AK')
# Change name of meetstate to region
colnames(num_meets)[1] <- 'region'
# Change abbreviation to name
num_meets$region <- state.name[match(num_meets$region, state.abb)] %>% tolower()

# Now normalize using state.x77
# Create the vector state.pop that includes population
state.pop <- state.x77[,1]
# Remove Hawaii and Alaska
state.pop <- state.pop[names(state.pop) != 'Alaska' & names(state.pop) != 'Hawaii']
# Make names lowercase
names(state.pop) <- names(state.pop) %>% tolower()
# Create a vector that maps state.pop and num_meets
match_vector <- match(names(state.pop), num_meets$region)
# Normalize
num_meets$NormNumMeets <- num_meets[match_vector,2] / state.pop

map + geom_map(data = num_meets, map = usa, aes(fill = NormNumMeets, map_id = region)) + 
   scale_fill_gradient(low = 'white', high = 'cyan4') + 
   theme(axis.title = element_blank(), panel.grid = element_blank(), 
         panel.background = element_blank(), axis.ticks = element_blank(), 
         axis.text = element_blank(), plot.title = element_text(hjust = .5, size = 16), 
         plot.subtitle = element_text(hjust = .5)) + 
   ggtitle('Number of Meets Per State', subtitle = 'Normalized by Population')
```

Minnesotta is a huge outlier, which skews the appearance of the rest of the map. 